---
title: "Homework_03"
author: "Akshaya Mahesh"
date: "10/21/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE,
                      fig.align = 'center',
                      out.width = '70%',
                      warning = FALSE,
                      message = FALSE)
```
### Installing packages
```{r eval=FALSE}
install.packages(c( "plyr", "dbplyr","usmap"))
```


### Adding the necessary libraries
```{r}
library(plyr)
library(tidyverse)
library(ggplot2)
library(readr)
library(dplyr)
library(usmap)
```
## Part 1
### Importing files and combining them
Dataset after binding rows
```{r }
files <- list.files(path = "C:\\Users\\maksh\\OneDrive\\Documents\\Datasets\\epa-aqi-data-annual\\epa-aqi-data-annual", pattern = "*.csv",
                    full.names = T)
data_all <- lapply(files, read.csv, header = TRUE, stringsAsFactors = FALSE)
air_aqi <- bind_rows(data_all)
head(air_aqi)
```

### Including plots to show Median AQI Values Vs Time
```{r}

ggplot(data=air_aqi,mapping = aes(y=Year,x=Median.AQI,group=Year))+geom_boxplot()+
  labs(title="Improvement in Air Quality Index over the years 1980-2019 ",
       x="Median AQI Values",y="Time in years")+theme_minimal()
```
Observation  for the above plot:  
1.The above plot shows that Air Quality Index in the United States has improved over the 40 year period.  
2.The range of spread of Median AQI values has decreased over the years.  
3.There is a decrease in the number of outliers having higher Median AQI values and increase in the number of outliers having very lower median AQI values.  

## Part 2
creating a new variable 'decade'    
```{r}
air_aqi$decade <-ifelse(air_aqi$Year>=1980 & air_aqi$Year<1990,
                        "1980-1989",
                        ifelse(air_aqi$Year>=1990 & air_aqi$Year<2000,
                        "1990-1999",
                        ifelse(air_aqi$Year>=2000 & air_aqi$Year<2010,
                        "2000-2009",
                        ifelse(air_aqi$Year>=2010 & air_aqi$Year<2020,"2010-2019","0"))))
head(air_aqi)

``` 
Selecting the State, Year,Decade and Median AQI columns for observation    
```{r}
air_aqi_decade<-air_aqi%>%select(c("State","Year","Median.AQI","decade"))
head(air_aqi_decade)
```

### Including plots for the Average Median AQI values across the states for four decades(1980-2019)

```{r message=FALSE}
my_aqi<-ddply(air_aqi_decade, .(decade,State),summarise,
              Average_MedianAQI=mean(`Median.AQI`,na.rm=TRUE))
my_aqi$region <- tolower(my_aqi$State)
head(my_aqi)
```
Using Fips to plot the data
```{r}

my_aqi$fips <- fips(my_aqi$region)

plot_usmap(data =my_aqi, values = "Average_MedianAQI", labels=FALSE)+
  scale_fill_viridis_c(option = "plasma")+theme(legend.position = "right")+
  theme(panel.background = element_rect(colour = "black")) +
  labs(title = "Decade wide Average AQI across the States")+
  facet_wrap(~decade)
  
```

Observation for the above plot:  
1. Air Quality Index seems to have improved over the decades across the United States.     
2. States such as Arizona, West Virginia and Pennsylvania show great improvement in Air Quality Index over the   period.    
3. On the contrary, Air Quality index in North Dakota,South Dakota, Arkansas and Hawaii has increased over the decades.    
4. Florida seems to have had a nearly constant Average AQI.    
5. In 1980-1989 , the states appear to have a wide range of Air AQI(comprises of both lower and extreme values) whereas in 2010-2019, the states appear to have a uniform AQI in the range of 30-50 (exception being California and Arizona).    

## Part 3
Importing the country and world_4region file  
```{r}
dir1 <- "C:\\Users\\maksh\\OneDrive\\Documents\\Datasets"
dir2 <- "ddf--gapminder--systema_globalis-master\\ddf--gapminder--systema_globalis-master"
path <- file.path(dir1, dir2,"ddf--entities--geo--country.csv" )
countries <- read_csv(path)
head(countries)
```

```{r}
path<-file.path(dir1, dir2, "ddf--entities--geo--world_4region.csv")
world_region <- read_csv(path)

head(world_region)
```
Using INNER JOIN to join the tables Countries and World_Region to get countries that have been assigned a world region  
```{r}
country_count<- inner_join(countries,world_region,by="world_4region")
head(country_count)
```
### Including Plot to visualize the number of countries across regions
```{r}
ggplot(data=country_count,mapping=aes(x=name.y,fill=name_short))+geom_bar()+
 labs(title="Total count of Countries in each World Region",x="Regions",
      y="Count (in numbers)",fill="WorldRegions")+
  geom_text(stat='count',aes(label=..count..),vjust=-0.2)+theme_minimal()
```
Observation of above plot:  
Asia seems to have the highest number of countries and The Americas the least.  

## Part 4
Importing the infant mortality rate datapoints table  
```{r}
dir3 <- "ddf--gapminder--systema_globalis-master\\ddf--gapminder--systema_globalis-master\\countries-etc-datapoints"
path<-file.path(dir1, dir3,"ddf--datapoints--infant_mortality_rate_per_1000_births--by--geo--time.csv")
inf_mortality_rate <- read_csv(path,guess_max=1000)
colnames(inf_mortality_rate)[colnames(inf_mortality_rate)=='geo']<-'country'
head(inf_mortality_rate)
```
Using INNER JOIN to join the country_count and inf_mortality_rate tables to get region wise infant mortality data  
```{r}
inf_region<-inner_join(inf_mortality_rate,country_count,by="country")
head(inf_region)
```
 
```{r}
colnames(inf_region)[which(names(inf_region)=="name.y")]<-"Region"
```
### Including plots for region-wise infant mortality rates 
```{r}
ggplot(data=inf_region,mapping=aes(x=time, y=infant_mortality_rate_per_1000_births,alpha=2)) +
     geom_point()+geom_smooth()+
  theme_minimal()+
labs(title = "Decrease in Infant mortality rates across the four regions",
     x="Time in Years",y="Infant Mortality Rate",
     subtitle="Region-wise Infant Mortality rates over two centuries")+
  facet_wrap(~Region)

```
Observation of the above plot:  
Overall infant mortality rate seems to have reduced over the given period across all the four regions.  

Africa: This region has had Infant mortality rate of 200-250 and has now come down to 5-100(approx).This range seems to be quite high when compared to other regions.  

Asia: This region seems to have had a nearly constant mortality rate for the period of 1850-1950 and then a slight increase during 1950s and then has started to reduce to 0-50(approx) at year 2000.  

Europe :Data for infant mortality for the 1800-1850 is available only for Europe out of the four regions. This region has had a nearly constant infant mortality rate of 200(approx.) for the years(1800-1900)and then starts to reduce constantly and now it seems to be 0-20(approx).  

The Americas: This region shows an increase in the rates for the period of 1850-1900 and the starts to decrease over the following years and at 2000, it seems to have an infant mortality rate of 0-50(approx.)  


## Part 5
Importing the life expectancy datapoints table  
```{r}
path<-file.path(dir1,dir3,"ddf--datapoints--life_expectancy_years--by--geo--time.csv")
life_exp<-read_csv(path)
colnames(life_exp)[colnames(life_exp)=='geo']<-'country'
head(life_exp)
```
Using INNER JOIN to combine the life_exp and inf_region table to help visualize Life Expectancy Vs Infant Mortality Rate  
```{r}
life_exp_Inf_rate<-inner_join(life_exp,inf_region,by=c("country","time"))
head(life_exp_Inf_rate)
```

```{r}
ggplot(data=life_exp_Inf_rate,mapping=aes(x=infant_mortality_rate_per_1000_births
                                          ,y=life_expectancy_years,alpha=2))+
  geom_point()+geom_smooth()+facet_wrap(~life_exp_Inf_rate$Region)+
  labs(title = "Plot to show relationship between Life Expectancy and 
       Infant mortality rate for regions",x="Infant Mortality Rate"
       ,y="Life Expectancy in Years",
       subtitle="Life Expectancy and  Infant Mortality rate seem 
       to vary negatively for all the regions")+
  theme_minimal()+facet_wrap(~Region)
```
Observation for the above plot:  
1.Life Expectancy and Infant Mortality rate seems to show negative correlation.  
2.For a period where Infant mortality rate is between 0-200, the data points are widely distributed for Africa and Asia whereas, Europe and The Americas seem to have a much lesser range.  
3.All the regions seem to have a life expectancy range of (30-80).  
